---
title: "Laplace approximation with xopt"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Laplace approximation with xopt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

This vignette shows how to use `xopt_laplace()` to compute the Laplace
approximation to an intractable integral of the form
$$Z = \int_{\mathbb{R}^p} \exp\bigl(-f(\theta)\bigr)\,\mathrm{d}\theta,$$
which is the ubiquitous shape of a (normalized or unnormalized) posterior
marginal likelihood in Bayesian inference.

The Laplace approximation expands $f$ to second order around its mode
$\theta^\star = \arg\min f$:
$$\log Z \;\approx\; -f(\theta^\star) + \tfrac{p}{2}\log(2\pi) - \tfrac{1}{2}\log\det H,$$
where $H = \nabla^2 f(\theta^\star)$. All three ingredients are handled by
xopt: `xopt_minimize()` locates the mode, a finite-difference Hessian
approximates $H$, and the new **differentiable** `xopt_logdet()` from
`xopt::linalg` computes $\log\det H$ with a reverse-mode-ready XAD tape.

## A Gaussian sanity check

For a target of the form $\exp(-\tfrac{1}{2}(\theta-\mu)^\top \Lambda (\theta-\mu))$
the Laplace approximation is exact: we should recover
$\log Z = \tfrac{p}{2}\log(2\pi) + \tfrac{1}{2}\log\det \Lambda^{-1}$.

```{r}
library(xopt)
set.seed(1)

p <- 4
L <- matrix(rnorm(p * p), p, p)
Prec <- crossprod(L) + diag(p)   # symmetric positive definite
mu   <- rnorm(p)

nlpost <- function(theta) {
  d <- theta - mu
  0.5 * as.numeric(t(d) %*% Prec %*% d)
}

res <- xopt_laplace(nlpost, par = rep(0, p))
res
```

Compare to the closed form:

```{r}
Sigma <- solve(Prec)
exact <- 0.5 * p * log(2 * pi) + 0.5 * as.numeric(determinant(Sigma, logarithm = TRUE)$modulus)
c(laplace = res$log_marginal, exact = exact, diff = res$log_marginal - exact)
```

The difference is at the level of finite-difference noise in the Hessian,
not the approximation itself (because the target is Gaussian).

## Bayesian logistic regression

A more realistic toy: `n = 50` i.i.d. binary observations with a 2D
covariate, standard-normal prior on the coefficients. We compare the
Laplace approximation of the marginal likelihood $\log p(y)$ across two
candidate priors with different prior precisions.

```{r}
set.seed(2)
n <- 50
X <- cbind(1, rnorm(n))                                    # intercept + 1 covariate
beta_true <- c(0.5, -1.2)
eta <- as.numeric(X %*% beta_true)
y <- rbinom(n, size = 1, prob = plogis(eta))

nlpost <- function(beta, tau) {
  eta <- as.numeric(X %*% beta)
  nll  <- -sum(y * eta - log1p(exp(eta)))                  # Bernoulli NLL
  nlp  <- 0.5 * tau * sum(beta^2)                          # N(0, 1/tau) prior
  nll + nlp
}

log_marginal <- function(tau) {
  xopt_laplace(function(b) nlpost(b, tau), par = c(0, 0))$log_marginal
}

taus <- c(0.1, 1, 10)
sapply(taus, log_marginal)
```

The prior with the best Laplace-approximated marginal likelihood is the
one the data prefer (here, `tau = 1`, which matches the N(0, 1) truth).

## What's differentiable, and why it matters

`xopt_logdet()` is backed by the same template-on-scalar implementation
that underlies every other `xopt::linalg` primitive. When the host code
records elementary operations onto an `xad::AReal<double>` tape,
`xopt_logdet()` joins the tape just like `+`, `*`, or `sqrt` would. This
means that in a future xopt release — once a C++-level compiled objective
exposes its own `Problem<Scalar>` to the Laplace glue — the derivative of
`log_marginal` with respect to hyperparameters (e.g., `tau` above) will
be available from a single reverse sweep, with no finite-difference loop
around `xopt_laplace()` needed.

From R today, the same idea already works if you supply an analytical
gradient and/or Hessian through `gr` and `hessian`:

```{r}
Prec0 <- diag(2)
hess  <- function(beta, tau) {
  p <- plogis(as.numeric(X %*% beta))
  W <- diag(p * (1 - p))
  t(X) %*% W %*% X + tau * Prec0
}

res <- xopt_laplace(
  fn = function(b) nlpost(b, tau = 1),
  par = c(0, 0),
  hessian = function(b) hess(b, tau = 1)
)
res$log_marginal
```

## See also

- `xopt_minimize()` for the full gradient-mode API (`"user"`, `"compiled"`,
  `"traced"`, `"fd"`, `"auto"`).
- `xopt_chol()`, `xopt_solve()`, `xopt_inv()` — the other three
  differentiable primitives `xopt_logdet()` is built on.
- `docs/ROADMAP_v3.md` for the planned implicit-differentiation API that
  will generalize `xopt_laplace()` to arbitrary fixed-point solves.
