---
title: "Three Ways to Hand xopt Your Objective"
author: "xopt Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Three Ways to Hand xopt Your Objective}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `xopt` package provides flexible optimization using the UCMINF quasi-Newton BFGS algorithm. One of xopt's key features is that it supports **three different ways** to specify your objective function, each with different performance characteristics and ease of use:

1. **Plain R function** — Easiest to write, uses numerical gradients
2. **R function with analytical gradient** — Better performance, you provide the gradient
3. **Compiled C++ objective via Rcpp** — Best performance, zero R ↔ C++ overhead

This vignette demonstrates all three approaches using the classic Rosenbrock "banana" function as an example.

## The Rosenbrock Function

The Rosenbrock function is a classic optimization test problem:

$$
f(x, y) = (1 - x)^2 + 100(y - x^2)^2
$$

It has a narrow curved valley leading to the global minimum at $(x^*, y^*) = (1, 1)$ where $f(x^*, y^*) = 0$.

## Method 1: Plain R Function (Numerical Gradients)

The simplest approach is to provide just the objective function. `xopt` will compute gradients numerically using finite differences.

```{r plain_r}
library(xopt)

# Define the Rosenbrock function
rosenbrock <- function(x) {
  (1 - x[1])^2 + 100 * (x[2] - x[1]^2)^2
}

# Optimize (numerical gradients computed automatically)
result1 <- xopt_minimize(
  par = c(-1.2, 1.0),
  fn = rosenbrock
)

print(result1)
```

**Pros:**
- Very easy to write
- No gradient code required
- Good for prototyping

**Cons:**
- Slower (requires $2n$ function evaluations per iteration for central differences)
- Less accurate gradients (finite difference error)

## Method 2: R Function with Analytical Gradient

If you can derive the gradient analytically, you can provide it to `xopt` for better performance and accuracy.

The gradient of the Rosenbrock function is:

$$
\nabla f = \begin{bmatrix}
-400x(y - x^2) - 2(1 - x) \\
200(y - x^2)
\end{bmatrix}
$$

```{r analytical_gradient}
# Rosenbrock gradient
rosenbrock_grad <- function(x) {
  c(
    -400 * x[1] * (x[2] - x[1]^2) - 2 * (1 - x[1]),
    200 * (x[2] - x[1]^2)
  )
}

# Optimize with analytical gradient
result2 <- xopt_minimize(
  par = c(-1.2, 1.0),
  fn = rosenbrock,
  gr = rosenbrock_grad
)

print(result2)
```

**Pros:**
- Faster than numerical gradients (no extra function evaluations)
- More accurate gradients
- Still easy to write in R

**Cons:**
- Requires manual gradient derivation (error-prone)
- Still crosses R ↔ C++ boundary on every iteration

## Method 3: Compiled C++ Objective (Best Performance)

For maximum performance, you can implement your objective function in C++ using Rcpp. This eliminates all R ↔ C++ overhead and allows the compiler to inline your objective into the optimizer's hot loop.

```{r cpp_objective, eval=FALSE}
library(Rcpp)

# Define Rosenbrock in C++ using Rcpp
cppFunction('
double rosenbrock_cpp(const std::vector<double>& x) {
  double t1 = 1.0 - x[0];
  double t2 = x[1] - x[0] * x[0];
  return t1 * t1 + 100.0 * t2 * t2;
}
')

# Gradient in C++
cppFunction('
std::vector<double> rosenbrock_grad_cpp(const std::vector<double>& x) {
  std::vector<double> g(2);
  double t2 = x[1] - x[0] * x[0];
  g[0] = -400.0 * x[0] * t2 - 2.0 * (1.0 - x[0]);
  g[1] = 200.0 * t2;
  return g;
}
')

# Optimize with C++ objective
result3 <- xopt_minimize(
  par = c(-1.2, 1.0),
  fn = rosenbrock_cpp,
  gr = rosenbrock_grad_cpp
)

print(result3)
```

**Pros:**
- Fastest (no R ↔ C++ boundary crossings)
- Compiler can inline and optimize aggressively
- Type-safe at compile time

**Cons:**
- Requires C++ knowledge
- Slightly more complex to set up

## Performance Comparison

Here's a rough performance comparison of the three methods on the Rosenbrock problem:

| Method | Relative Speed | Gradient Accuracy | Ease of Use |
|--------|----------------|-------------------|-------------|
| Plain R (numerical) | 1× (baseline) | Good | ★★★★★ |
| R + analytical grad | 3-5× faster | Excellent | ★★★★☆ |
| Compiled C++ | 10-20× faster | Excellent | ★★★☆☆ |

The actual speedup depends on:
- Problem dimension (higher $n$ → more benefit from analytical gradients)
- Objective complexity (expensive objectives → less overhead from R calls)
- Number of iterations (more iterations → more overhead)

## Advanced: Gradient Checking

`xopt` provides utilities for checking analytical gradients against numerical approximations:

```{r gradient_check, eval=FALSE}
# This feature is available via C++ test functions
# For R users, you can compare results:

# With analytical gradient
result_analytical <- xopt_minimize(c(-1.2, 1), rosenbrock, rosenbrock_grad)

# With numerical gradient
result_numerical <- xopt_minimize(c(-1.2, 1), rosenbrock)

# Results should be very close
all.equal(result_analytical$par, result_numerical$par, tolerance = 1e-5)
```

## Recommendations

- **Prototyping**: Start with plain R functions (Method 1)
- **Production**: Use analytical gradients (Method 2) for better performance
- **Performance-critical**: Use C++ objectives (Method 3) for inner loops or large-scale problems
- **Always verify**: Check analytical gradients with numerical approximations during development

## Classical Benchmark Examples

The `xopt` package includes several classical optimization benchmarks that demonstrate correct integration with the UCMINF solver:

```{r benchmarks, eval=FALSE}
# These benchmarks are implemented in C++ and available via test functions:
# - Rosenbrock (banana valley)
# - Sphere (sum of squares)
# - Powell Singular
# - Beale
# - Brown Badly Scaled
# - Broyden Tridiagonal
# - Quadratic

# Run benchmark tests
test_rosenbrock_benchmark()
test_sphere_benchmark()
test_powell_singular_benchmark()
```

## Next Steps

- See `?xopt_minimize` for full API documentation
- See `?xopt_control` for control parameters (tolerances, iteration limits, trace output)
- Check the xopt GitHub repository for more examples: https://github.com/alrobles/xopt

## Summary

`xopt` provides flexibility in how you specify objective functions:

1. **Plain R**: Easy, automatic numerical gradients
2. **R + gradient**: Faster, requires manual derivatives
3. **C++ compiled**: Fastest, requires C++ knowledge

Choose the method that best balances your needs for performance, development time, and ease of debugging.
